Predictive modeling using genomic annotations

### 2018-10-03 Manuscript/manuscript.Rmd
edits

---
title: "Predictive modeling using genomic annotations"
subtitle: "Pitfalls and Recommendations"
author: "Spiro Stilianoudakis"
date: "August 9, 2018"
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{amsmath}
output: 
  word_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 2
    toc: yes
  pdf_document:
    toc: yes
---

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 cell line (Rao et. al).  The domain data was constructed from in situ Hi-C contact matrices at a 5kb resolution. Identified TAD boundaries were represented by their location in the genome (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

@@

### Question 1: How severely unbalanced were the classes and what was the affect on model performance?

When binning the genome, the sparcity of TAD boundaries contributed to heavily imbalanced classes. That is, proportionally, only a small number of genomic bins overlapped with a flanked TAD boundary. To assess this issue two multiple logistic regression models were performed. One with unbalanced classes and one with balanced classes created by random under-sampling from the majority class. It was found that the model with balanced classes out-performed the model with unbalanced classes. However, given the vast disparity between classes simply performing one iteration of random under-sampling was not sufficient in determining the best possible balancing technique.

### Question 2: Which balancing technique performs best?

Three methods were used to evaluate class balance. The first consisted of performing X iterations of over-sampling from the minority class to match the majority class. At each iteration, a multiple logistic regression model was used and performance metrics were recorded. At the end of the X iterations, the performance metrics were aggregated by taking the average. The second method consisted of random under-sampling. Similar to over-sampling, the majority class was randomly under-sampled over X iterations to create balanced classes. Again, at each step a multiple logistic regression model was performed and the results were aggregated using the average.  The last method incorporated the SMOTE command from the DmWR package in R. SMOTE stands for Synthetic Minority Over-sampling Technique. The SMOTE algorithm incorporates both under-sampling (percent under) of the majority class and over-sampling (percent over) of the minority class. The minority class is over-sampled by taking each minority class sample and introducing synthetic examples along the line segments joining any/all of the k minority class nearest neighbors (here, k=5). Then, the majority class is under sampled by taking a random sample that matches some pre-specified percentage of the updated minority class. A total of four different combinations of percent over and under-sampling were used which included: 100/200, 200/200, 300/200, and 400/200. It is important to note that using a combination of 100/200 (percent over/percent under-sample) will always result in perfectly balanced classes. The other combinations created unbalanced classes with the negative class as the majority to reflect the proportionality of the classes from the original data. Multiple logistic regression models were used and performance metrics were assessed and compared with each of the other techniques. It was determined that the model with data untilizing random under-sampling out-performed all other other balancing techniques. However, to this point we only considered overlaps between genomic bins and the regions that defined functional genomic elements.

### Question 3: What affect does the inclusion of distances from bin to region have on model performance?

To this point we had only considered overlaps between genomic bins and the regions that defined functional genomic elements. Now we added another type of predictor created from each genomic element. This consisted of the distance from the center of each genomic bin to the center of the nearest region defining a genomic element as depicted in Figure X. To assess any improvement with the inclusion of these predictors we compared the performances of three multiple logistic regression models: one with only the overlap predictors, one with only the aforementioned distance predictors, and one with both.  It was found that the inclusion of distance predictors improved model performance.  Likewise, the model with both types of predictors futher improved model performance. As is common, with genomic data, the use of continuous predictors (especially with regards to distances) results in highly right skewed data with distances ranging from 0 to tens of thousands in base pairs. This can affect the convergence of likelihood maximization.

### Question 4: What normalization technique performed on the distance predictors best improves model performance?

Here, we considered three normalization techniques. The first was a simple $log_{2}$ transformation, in which one was added in the case of distances of 0. The second was an inverse hyperbolic sine tranformation given by $insert function here$ which has the benefit of handling values of 0. Lastly, an inverse normal tranformation was considered. The inverse normal transformation is a way of transforming the sample distribution of a continuous variable to make it appear more normally distributed by matching the normalized rank of the trait to a quantile in a normal distribution using the following formula $insert formula here$. Multiple logistic regression models were performed using data with each respective transformation and compared. It was found that the model with inverse normally transformed distance predictors not only out-performed the other transformation but also out-performed a model with no transformation at all. So far the only classification algorithm considered was a multiple logistic regression. However, given the large number possible functional genomic elements that can be considered, this model can be computationally inefficient.

### Question 5: Which classification algorithm most accurately and efficiently identifies the genomic elements associated with TAD boundaries?

We compared the results from a multiple logistic regression model to that of a multiple logistic regression model with both LASSO and Elastic-Net regularization. Prior to model fitting, the predictors were standardized in order to compare magnitudes of the subsequent beta coefficients as a measure of variable importance. For the models with regularization, 10-fold cross validation was used to tune the penalization terms ($\lambda$ for LASSO and both $\lambda$ and $\alpha$ for Elastic-Net). Model performance was compared using F1-scores, MCCs, and AUROCs. The alpha level of significance was set at 0.05. All statistical analyses was performed in R version 3.4.2.

@@

There was a total of 247632 genomic bins that made up the response vector Y, 1629 (0.7%) of which contained TAD boundaries (Y=1). A total of 68 genomic features were included in the analysis. Two predictors were included for every feature (one binary, one continuous), making for a total of 136 predictors.  The list of predictors is provided in Table X. The feature space was reduced to 90 after all near-zero variance predictors were removed. The predictors that were removed are listed in Table X [??? Mark them in supplementary table of all features].

The best normalization technique was found to be the model that included a log base 2 transform with no standardized predictors. The AUC was found to be 0.809 and presented in Figure X. We see that this value is only marginally greater than the model using a log base 2 transform and standardized predictors. However, from Figure X, we see that the variable importance plot associated with a log base 2 transform with no standardized predictors (top right) yielded more accurate results than all other combinations. Notably, the insulator, CTCF, and DNaseI features are located in the top 3 most important features for predicting TAD boundaries, as expected. This is in contrast to the other three combinations, which presented conflicting results. Thus, it was concluded that only performing a log base 2 transformation was necessary.

Figure X presents the model performance metrics of the eight different SMOTE combinations. The SMOTE combination utilizing percent over of 100 and percent under of 200 yielded the largest AUC (0.794). This was to be expected since this combination created perfectly balanced classes. However, looking at Figure X, the method using 100 bootstrap sample performed better than then model using SMOTE with an AUC of 0.809. Thus, it was concluded that using bootstrap samples would be more efficient moving forward.

The best performing variable selection technique was found to be forward selection (AUC: 0.810) as shown in Figure X. All of the stepwise procedures performed better than the recursive feature elimination method, which not only performed worse, but also chose the most predictors. It should be noted that both forward and backward selection had the same AUC. However, the forward selection chose fewer predictors, 28 compared to 34 respectively. Thus, forward selection was chose as the most appropriate variable selection technique. Figure X presents the relationship between each variable selection technique. We see that there was significant agreement between forward and backward selections, with 26 predictors being shared by both. Likewise, there was a total of 12 predictors in common between all four techniques. Table X lists the common predictors between each comparison.

After the data was ran through our pipeline, the final dataset consisted of 28 log transformed predictors. Lastly, 100 bootstrap sampled random forest models were performed on said data. Figure X presents the results compared to the multiple logistic regression models proposed by Mourad et al. We see that the random forest model, using the pipeline processed data, preforms better compared to both models proposed by Mourad et al with an AUC of 0.805. Moreover, the variable importance plot in Figure X reiterates the strong predictive performance of the random forest model by indicating that the DNaseI, insulator, and CTCF genomic features are among the top 3 variables most predictive TAD boundaries. The MLR with LASSO estimation only marginally out performs the regular MLR model with AUCs of 0.750 and 0.747 respectively. Likewise, from Table X we can see that there is more similarity in the ranking of important features between the random forest model and the MLR with LASSO model. However, there is still a large disparity, with 12 variables recognized as important by the random forest that are missing from the MLR with LASSO model. Thus, we conclude that a random forest algorithm applied to a dataset ran through the proposed pipeline performs uniformly better than a multiple logistic regression model, regardless if LASSO estimation is used.

>>>>

---
title: "Predictive modeling using genomic annotations"
subtitle: "Pitfalls and Recommendations"
author: "Spiro Stilianoudakis"
date: "August 9, 2018"
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{amsmath}
output: 
  pdf_document:
    toc: yes
  word_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 2
    toc: yes
---

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 cell line (Rao et. al).  The domain data was constructed from in situ Hi-C contact matrices at a 5kb resolution. Identified TAD boundaries were represented by their location in the genome (hg19 human genome assembly), including chromosome, start and end coordinates for chromosomes 1 through 22. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

@@

### Class Imbalance

When binning the genome, the sparcity of TAD boundaries contributed to heavily imbalanced classes. That is, proportionally, only a small number of genomic bins overlapped with a flanked TAD boundary. To assess this issue a series of multiple logistic regression models were performed. First, a multiple logistic regression model using the full data with unbalanced classes was assessed. We then compared this to a multiple logistic regression model with balanced classes created by random over-sampling, random under-sampling, and four other multiple logistic regression models with balanced classes created by the SMOTE (Synthetic Minority Over-sampling Technique) function in the DmWR package in R. For each of the random over- and under-sampling methods, multiple iterations were considered up until the difference in the cumulative means reached below a threshold of $5 \times 10^{-5}$. Performances were then aggregated by taking the average across all iterations. It was determined that the model with data untilizing random under-sampling out-performed all other other balancing techniques. However, to this point we only considered overlaps between genomic bins and the regions that defined functional genomic elements.

### Incorporating Distances

To gain further insight into the spatial relationship between these chromatin factors and the genome, we introduced distance predictors into the model. This consisted of the distance from the center of each genomic bin to the center of the nearest region defining a genomic element as depicted in Figure X. To assess any improvement with the inclusion of these predictors we compared the performances of three multiple logistic regression models: one with only the overlap predictors, one with only the aforementioned distance predictors, and one with both.  It was found that the inclusion of distance predictors improved model performance.  Likewise, the model with both types of predictors futher improved model performance. As is common, with genomic data, the use of the distance predictors resulted in highly right skewed data with distances ranging from 0 to tens of thousands in base pairs.

To better improve computational efficiency with regards to the convergence of likelihood maximization, we considered three normalization techniques. The first was a simple $log_{2}$ transformation, in which one was added in the case of distances of 0. The second was an inverse hyperbolic sine tranformation given by $insert function here$ which has the benefit of handling values of 0. Lastly, an inverse normal tranformation was considered. The inverse normal transformation is a way of transforming the sample distribution of a continuous variable to make it appear more normally distributed by matching the normalized rank of the trait to a quantile in a normal distribution using the following formula $insert formula here$. Multiple logistic regression models were performed using data with each respective transformation and compared. It was found that the model with inverse normally transformed distance predictors not only out-performed the other transformation but also out-performed a model with no transformation at all.

### Classification Models

Therefore, the final data used for quantifying the association between each functional genomic element and TAD boundary formation was one that had been balanced using multiple iterations of random under-sampling, incorporated both overlapped and distance predictors, and normalized using an inverse normal transformation. Likewise, prior to model fitting, the predictors were standardized in order to compare magnitudes of the subsequent beta coefficients as a measure of variable importance.  We then performed a wide variety of classification algorithms to determine which genomic elements were most predictive of TAD boundaries. The models included a multiple logistic regression, a multiple logistic regression with both LASSO and Elastic-Net regularization, a random forest, a gradient boosting machine, a support vector machine, and a naiive bayes model. For each model, 10-fold cross validation was performed to tune for any neccessary paramenters. Model performance was compared using F1-scores, MCCs, and AUROCs. The alpha level of significance was set at 0.05. All statistical analyses was performed in R version 3.4.2.

@@

A total of 9048 topologically associating domains and their genomic locations were obtained for the GM12878 cell line for chromosomes 1 through 22.  The median width of TADs was found to be 18.5 kb (IQR: 11.5kb - 33.0kb). The start and end coordiantes of the TADs were translated into a vector of boundary points. There were 1454 (16%) TADs that shared at least one coordinate (start, end, or both) with another TAD. In this case the unique boundary point demarcating two adjacent TADs was considered. The genome was binned at 10 kb interval, resulting in a total of 536396 bins. A total of 536396 genomic bins of 10 kb width were created. After flanking the TAD boundary points on either side by 5 kb, it was found that 31820 (5.9%) bins contained a flanked TAD boundary. We investigated a total of 68 genomic features, 12 (18%) of which were removed because they were assessed as having near-zero variance. Table X presents a full list of the genomic features in addition to the metrics and indicator corresponding to whether or not the feature was to be removed.

## Balancing Classes Improves Performance

It was found that the multiple logistic regression models utilizing random over and under-sampling out-performed the other techniques and were comparatively similar with F1 scores of 0.636 and 0.637 respectively. The under-technique was evaluated over 13 iterations, while the over-sampling technique took far more iterations for the cumulative mean F1 score to converge, X to be exact. Given that and the fact that the under-sampling technique uses significantly less data than the over-sampling, while still performing at the same level, it was concluded that this was the best balancing technique and used in further analyses.

## Incorporating Distance Predictors Improves Performance

The multiple logistic regression model with only distance predictors, evaluated on multiple iterations of random under-sampling was found to improve model performance from 0.637 (using only overlap predictors) to 0.747. Furthermore, the multiple logistic regression model with both types of predictors further improved model performance by raising the F1 score to 0.753. This suggested that the distance predictors were more representative of the association between the genomic elements and TAD boundaries. However, on a genomic scale, distances from the bins to the nearest region defined by a respective genomic element were highly right skewed. We evaluated three tranformations in order to normalize the distance predictors and improve model performance. Out of the three normalization techniques, the multiple logistic regression model with an inverse normal transformation on the distance predictors was found to perform the best with an F1 score of 0.747.

##
